---
title: "Linear models"
author: "Mark Andrews"
date: "July 5, 2018"
output:
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

# Introduction

In linear models, we model the expected value of an outcome variable as a linear function of one or more predictor variables. Even when we know that this is not a great modelling assumption, linear models can still be very informative, especially for exploratory work. In any case, it is hard to progress to more complex and realistic models without first understanding linear models.

For the following, we will use some data from the ```psych``` package. So, first load that, and a few other goodies:
```{r message=FALSE}
library(MASS)
library(car)
library(psych)
library(dplyr)
library(ggplot2)
library(pander) # for making nice tables

Df <- mutate(sat.act, 
             gender = factor(gender))
```

We'll start by predicting *ACT* (a standardized academic test) scores on the basis of education level (measured on a five point scale):

```{r}
M <- lm(ACT ~ education, data=Df)
pander(summary(M))
```

We can visualize this as follows:
```{r}
ggplot(Df, 
       aes(x= education, y=ACT)) +
  geom_point() + 
  stat_smooth(method='lm')
```

## Confidence intervals
We can get confidence intervals as follows:
```{r}
confint(M)
```


## Predictions
On the basis of our fitted model ```M```, we can make predictions about possible values of the predictor variable. 
```{r}
hypothetical.data <- data.frame(education = c(1, 2, 5, 10, 15))
predict(M, newdata=hypothetical.data)
```
\pagebreak


# Multiple linear regression
We can add as many predictor variables as we like:
```{r}
M <- lm(ACT ~ education + age + gender, data=Df)
pander(summary(M))
```

## Collinearity
We'll evaluate multicollinerity using Variance Inflation Factor (VIF):
```{r}
vif(M)
```

# General linear models
We can use predictors that categorical as well as continuous in our model. Here, we investigate how the post treatment weight of a patient differs from their pre treatment weight, for three different types of therapy (control, CBT, family therapy).

First, we'll visualize the data (we'll turn off standard error shading to allow the lines to be seen more easily):
```{r}
ggplot(anorexia,
       aes(x = Prewt, y = Postwt, col=Treat)) +
  geom_point() +
  stat_smooth(method='lm', se=F) +
  theme_classic()
```

Now, we'll do a *varying intercept*, which is also known as an *ANCOVA*:
```{r}
M <- lm(Postwt ~ Prewt + Treat, data=anorexia)
pander(summary(M))
```

We also do a *varying slopes and varying intercepts* model. This is a type of interaction model:
```{r}
M_interaction <- lm(Postwt ~ Prewt * Treat, data=anorexia)
pander(summary(M_interaction))
```

## Model evaluation

We can compare any two linear models using the generic ```anova``` function. Here, we'll use this to test whether the varying slopes and intercepts model is a better fit to the data than the just varying intercepts model:

```{r}
model_comparison <- anova(M, M_interaction)
pander(model_comparison, missing='')
```


