---
title: "Simple plots"
author: "Mark Andrews"
date: "April 5, 2017"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---

# Introduction

There are two main ways to do plots in R: base R plots, and ggplot plots. There are other approaches too, but these two are the most widely used. Here, we cover base R plots.

Base R plots allows you to do simple plot relatively easily, but begins to be unwieldy when you need to more complex things. On the other hand, there is a lot of fine grained control in base R plots, so with persistence, you can usually get what you need.

# Histograms

For this, we'll get a data frame, and do a histogram of the log of the latency (reaction time).
```{r}
Df <- read.csv('../data/LexicalDecision.csv', header=T)
hist(log(Df$latency))
```

We can use more bins than the default of 10.
```{r}
hist(log(Df$latency), 25)
```

Now, we'll give it a better title, a better set of labels and fix the axes:

```{r fig.height=4}
hist(log(Df$latency), 
     xlab='Log of Latency',
     ylab='Frequency',
     xlim=c(5.5, 8),
     main='My prettier histogram',
     breaks=25)
```

# Density plot

We could also do a density plot:
```{r fig.height=3.5}
d <- density(log(Df$latency)) # get the density data
plot(d) # plots the results 
```

And we'll fix that up too
```{r fig.height=3.5}
plot(d,
     xlab='Log of Latency',
     ylab='Density',
     xlim=c(5.5, 8),
     main='Density plot of (log) reaction times')
polygon(d, col="gray", border="black") 
```

# Scatterplots

We can make scatterplots with the *plot* function and the formula ```y ~ x```, where `y` is the outcome variable (y axis variable) and `x` is the predictor variable (x axis variable).

```{r fig.height=4}
# Plot latency as a function of log frequency
plot(latency ~ log(frequency), data=Df)
```

That's a big ugly, so we'll fix it up:
```{r}
plot(latency ~ log(frequency), 
     xlim=c(-1, 6),
     ylim=c(250, 1500),
     xlab='Log of word frequency',
     ylab="Reaction time (seconds)",
     main='Predicting reaction times',
     pch=16, # point character
     cex=0.5,
     col='blue',
     data=Df)
abline(lm(latency ~ log(frequency), data=Df), col='red', lwd=3)
```
```

