---
title: "Psychology's Core Statistics Curriculum (More or less)"
output: 
  pdf_document:
    keep_tex: no
    pandoc_args: [
      "-V", "classoption=twocolumn"
    ]
geometry: top=1cm, bottom=2cm, left=1cm, right=1cm
tables: true
---

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```

```{r setup, include=FALSE}
set.seed(147001)
knitr::opts_chunk$set(echo = TRUE)
library(xtable)
library(knitr)
library(schoRsch)
```

## Read in a data file
```{r cars}
Df <- read.csv('../data/memorytest.csv')
```
## Descriptive statistics

Now, we'll get the mean, median, maximum, minimum, standard deviation, sample size, of the memory scores for both conditions.
```{r}
descriptives <- Df %>% group_by(condition) %>%
  summarize(mean = mean(memory),
            median = median(memory),
            max = max(memory),
            min = min(memory),
            sd = sd(memory),
            size = n())
```
```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(descriptives)
```

## Independent samples t-test
An independent samples t-test with the homogeneity of variances assumption is as follows:
```{r}
results <- t.test(memory ~ condition, data=Df, 
                  var.equal=T)
```
```{r, results='asis', echo=FALSE}

options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(t_out(results, print=FALSE))
```
## Paired samples t-test
```{r, echo=F}
N <- 25
```

For the paired samples t-test, we'll make up some data. We'll make two paired samples of normally distributed random numbers of size $N=`r N`$.
```{r}
x <- rnorm(N)
y <- rnorm(N)
results <- t.test(x,y, paired=T)
```
```{r, results='asis', echo=FALSE}

options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(t_out(results, print=FALSE))
```


## Z-scores
<!--We calculate z scores my subtracting the mean from a set of values and dividing by their standard deviation.-->
```{r}
scores <- rnorm(N, mean=100, sd=15)
z.scores <- (scores - mean(scores)) / sd(scores)
```

We get the area under the normal curve below each z-score as follows:
```{r}
area.under.curve <- pnorm(z.scores)
```

## $\chi^2$ test

We'll make a $2 \times 2$ contingency table:
```{r}
X <- matrix(c(120, 57, 75, 17), ncol = 2)
```
A $\chi^2$ test (with no continuity correction) is as follows:
```{r}
results <- chisq.test(X, correct=FALSE)
```

```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(chi_out(results, print=FALSE))
```

## Simple linear regression

First,  we'll read in our data from another csv file.
```{r}
Df <- read.csv('../data/chocolate.csv')
```

Now, we'll do a scatter plot and line of best fit. 
```{r}
qplot(chocolate, happiness, data=Df) + 
  geom_smooth(method='lm')
```

The regression model is fit with 
```{r}
M <- lm(happiness~chocolate, data=Df)
```
```{r echo=F}
S <- summary(M)
r.sq <- round(S$r.squared, 2)
adj.r.sq <- round(S$adj.r.squared, 2)
F.stat <- S$fstatistic
df.1 <- F.stat['numdf']
df.2 <- F.stat['dendf']
F.stat <- round(F.stat['value'], 2)
```

From this, we have $R^2=`r r.sq`$ (adjusted $R^2=`r adj.r.sq`$), $F(`r df.1`, `r df.2`) = `r F.stat`$. The coefficients table is as follows:
```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(summary(M))
```

## One-way ANOVA
Again here, we'll read in a new data set from a csv file. 

```{r}
Df <- read.csv('../data/antidepressants.csv')
```

For this, we coul use the `lm` (linear model) command, but instead we'll use `aov`. 
```{r}
M <- aov(depression ~ drug, data=Df)
```

```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(summary(M))
```

## Mann Whitney U test:

```{r}
x <- rnorm(10)
y <- rnorm(12)
results <- wilcox.test(x, y)
```


The U statistic is $U = `r results$statistic`$, $p=`r round(results$p.value, 2)`$.

## Wilcoxon signed ranks test:
```{r}
N <- 10
x <- rnorm(N)
y <- rnorm(N)
results <- wilcox.test(x, y, paired = TRUE) 
```

The W statistic is $W = `r results$statistic`$, $p=`r round(results$p.value, 2)`$.

## Multiple linear regression

```{r}
Df <- read.csv('../data/stress.csv')
M <- lm(stress ~ social + extra + perfect,
        data=Df)
```

```{r echo=F}
S <- summary(M)
r.sq <- round(S$r.squared, 2)
adj.r.sq <- round(S$adj.r.squared, 3)
F.stat <- S$fstatistic
df.1 <- F.stat['numdf']
df.2 <- F.stat['dendf']
F.stat <- round(F.stat['value'], 2)
```

From this, we have $R^2=`r r.sq`$ (adjusted $R^2=`r adj.r.sq`$), $F(`r df.1`, `r df.2`) = `r F.stat`$. The coefficients table is as follows:
```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(summary(M))
```

## Factorial Anova 

Here, we look at the effect of two variables (drug type and therapy type) and their interaction on depression scores.
```{r}
Df <- read.csv('../data/drugtherapy.csv')
M <- aov(depression ~ drug*therapy, data=Df)
```


```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(summary(M))
```

## Repeated Measures Two Way Anova

```{r}
Df <- read.csv('../data/movies.csv')
M <- aov(stress ~ music * image +
           Error(PID / (music * image)),
         data=Df)
```


```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(M)
```

## Repeated Measures Three Way Anova (One Between, Two Within) Anova 
```{r}
Df <- read.csv('../data/movies2.csv')
M <- aov(stress ~ time * music * image + 
           Error(PID / (music * image)),
         data=Df)
```
```{r, results='asis', echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(M)
```


