---
title: "Introducing vectors"
author: "Mark Andrews"
date: "April 5, 2017"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---
# Introduction

Vectors are simply arrays or lists of numbers. Most of time, when we are working with data, we work with *data frames*. Data frames can be seen as similar to spreadsheets, i.e. with multiple rows and multiple columns, and each column representing a variable. Here, each column is a vector and often we need to work directly with it.

We'll start by creating a vector of random numbers by using the `runif()` function. 

```{r}
set.seed(10101) # re-set the random number generator (for reproducibility)
(x <- runif(100)) # get a set of 100 random numbers, each between 0 and 1
```

# Let's examine the vector

```{r}
class(x) # What kind of object is it?
length(x)
str(x)   # compactly display internal structure of x
summary(x) # summarize the info in x
```

# Indexing 
* What is the value of element 68?
```{r}
x[68]
```

# Slicing
Slicing will give us a contiguous subset of the vector. For example, what are the values of elements 12 to 18 inclusive? 
```{r}
x[12:18]
```

# Multiple indices and subsets

What are the values of elements 17, 89, 39, 42? To do this, first create a vector of those elements and call it *indx* and then use this to slice *x* 
```{r}
indx <- c(17, 89, 39, 42)
x[indx]
```
This is the same thing as doing 
```{r}
x[c(17, 89, 39, 42)]
```

What if we needed to find all elements that are equal to or less than the value of 0.2? Here, we can create an indexing vector called *indx* and then use this to extract the elements:
```{r}
(indx <- which(x <= 0.2)) # Get indices of elements whose values are = or < 12.
x[indx]
```
We also use Boolean indices here:
```{r}
(indx <- x <= 0.2)
x[indx]
```

Note that to find how many elements are less than or equal to e.g. 0.2, we can do
```{r}
indx <- x == 0.2
sum(indx) # Sums up the Boolean vector 
```
which gives the same result as 
```{r}
indx <- which(x == 0.2)
length(indx)
```

**Warning**: Remember that *<=* is an inequality test, and *<-* is an assignment operator, and *==* is an equality test and *=* is an assignment operator.


## Descriptive statistics

We can easily get things like *mean*, *median*, *sd*, etc, etc.

```{r}
mean(x)
median(x)
sd(x)
var(x)
min(x)
max(x)
range(x)
IQR(x) # inter quartile range
```

This will give us some standard percentiles,
```{r}
quantile(x)
```
and we can ask for specific percentiles too:
```{r}
quantile(x, probs = c(0.025, 0.25, 0.5, 0.75, 0.975))
```


## Concatenating vectors

We can join up vectors using the generic "combine" function *c()*:
```{r}
z <- c(0.1, 0.2, 0.3, 0.27, 0.42)
(y <- c(x, z))
```

