---
title: "Data visualization"
author: "Mark Andrews"
date: "July 9, 2017"
output:
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: inline
---
# Introduction

R provides us with many tools for doing plotting and visualization. The dominant modern way is to use a program called
```ggplot```, which is essentially a mini-language for data visualization. 

```{r}
library(readr)
library(ggplot2)

```

We'll read in our data
```{r}
Df <- read_csv('../data/LexicalDecision.csv')
```

# A simple scatter plot

We use `ggplot` by first specifying the data frame to use and then which variables should be assigned to the x and y axes. Then we specify that we want points for the values.  
```{r fig.height=3}
ggplot(Df, aes(x=log(frequency), y=latency)) + geom_point()
```

We can add another *layer* that is a line of best fit, plus standard error:
```{r fig.height=3}
ggplot(Df, aes(x=log(frequency), y=latency)) + 
  geom_point() +
  stat_smooth(method='lm')
```

We can play with the properties of all the features of our plot:
```{r fig.height=3}
ggplot(Df, aes(x=log(frequency), y=latency)) + 
  geom_point(size=1, shape=1, colour='blue') +
  stat_smooth(method='lm') + 
  coord_cartesian(xlim = c(0, 5), ylim=c(250, 1000)) 
```

# Colour and shape codes

One of the great things about `ggplot` is how it allows us to colour code our data. For example, here we make a scatter plot and color code the points that belong to males and females. (When using alternative plotting methods to `ggplot`, doing things like this become very cumbersome):
```{r}
# Load a new data set
load('../data/beautyeval.Rda')

# Make scatterplot, point colour by male/female
ggplot(beautydata, aes(x=beauty, y=eval, colour=sex)) + 
  geom_point(size=1) +
  scale_color_manual(values=c('blue', 'red')) +
  scale_y_continuous(name='Teaching evaluation score') +
  scale_x_continuous((name='Attractiveness rating'))
```

As before, we can superimpose lines of best fit, etc.

```{r}
# Scatterplot, with lines of best fit and errors
ggplot(beautydata, aes(x=beauty, y=eval, colour=sex)) + 
  stat_smooth(method='lm') + 
  geom_point(size=1) +
  scale_color_manual(values=c('blue', 'red')) +
  scale_y_continuous(name='Teaching evaluation score') +
  scale_x_continuous((name='Attractiveness rating')) 
```

# Facets 

Facets allow us to create multiple plots in the same way, with each one showing some subset of the data.

```{r fig.height=10}
# Scatterplot, with lines of best fit and errors
# one per tenure group
ggplot(beautydata, aes(x=beauty, y=eval, colour=sex)) + 
  stat_smooth(method='lm') + 
  geom_point(size=1) +
  scale_color_manual(values=c('blue', 'red')) +
  scale_y_continuous(name='Teaching evaluation score') +
  scale_x_continuous((name='Attractiveness rating')) +
  facet_grid(tenure ~ .)
```




